A Coq Formalization of Stable Sorts, Radix Sort, and the Burrows-Wheeler Transform

### 2019-10-18 draft_paper/draft_paper.tex
Finish revising appendices

Stable sorts are an important and fundamental type of sorting
algorithm. A sort is stable if it preserves the relative order of
equal elements. For example, if we are stably sorting the list
$[\var{mix},\ \allowbreak\var{fog},\ \allowbreak\var{age},\ \allowbreak\var{fit}]$
by the first character of each word, we have to produce
$[\var{age},\ \allowbreak\var{fog},\ \allowbreak\var{fit},\ \allowbreak\var{mix}]$
and not
$[\var{age},\ \allowbreak\var{fit},\ \allowbreak\var{fog},\ \allowbreak\var{mix}]$.
Even though both are sorted on the first character, because \var{fog}
preceded \var{fit} in the original list, it must also precede
\var{fit} in the stably sorted list.

Stability can be directly useful in certain contexts, such as when
sorting a table by a single column. However, stability is also crucial
to the correctness of certain general-purpose algorithms. One such
algorithm, or set of algorithms, is radix sort. Radix sorts are used
to sort data lexicographically, and can be more efficient than
comparison-based sorting methods, both in theory and in practice
\cite{McIlroy93,Bentley:1997:FAS:314161.314321,10.1007/978-3-540-89097-3_3}.
Most radix sorts rely on a some stable sorting subroutine.

@@

Throughout our proofs about LSD radix sort and the BWT, we reason
about sorting algorithms whose implementation is irrelevent as long as
they are correct. We could, in principle, rely only on the fact that
they fulfill the specification of a sorting algorithm
(\cref{def:sort}). However, in practice this is difficult, because the
inductive definition of permutations can produce unhelpful induction
hypotheses, especially in the transitive case. Insetad, we reason
about specifically insertion sort. Therefore
\cref{thm:stablesort_unique} is quite useful, because we can reason
using insertion sort, and later on replace it with any other
(presumably faster) stable sort.

@@

Whenever we sort a list, we do so with respect to some ordering on its
elements. For example, we could sort a list of students by first name,
last name, birthday, etc. More precisely, for sorting we need a total,
decideable preorder: a \textit{preorder} is a relation $\le$ that is
reflexive and transitive, it is \textit{total} if we can compare any
two elements, and it is \textit{decidable} if we have a procedure that
takes any two elements $x$ and $y$ and tells us if $x \le y$ or $x \ge y$.

Given a total decideable preorder, we can define when a list is in
ascending order with respect to that preorder. In this case, as will
be the case with other properties we will discuss, there are multiple
reasonable definitions. Our approach will be to state all the
definitions we can think of and to prove equivalence between them.
This increases our confidence that we have defined things correctly,
and allows us to choose whichever definition we find most convenient
for a given proof.

@@

However, a sorting function cannot produce just any sorted list; the
output must also contain the same elements as the original list. To
capture this, we could use the \var{Permutation} relation defined in
the Coq standard library. This relation holds on two lists when they
contain all the same elements, and is defined inductively:
\begin{definition}[Permutation]
  \label{def:permutation}
  \begin{gather*}
    \infer[\var{perm\_nil}]{\var{Permutation}~[]~[]}{}
    \\
    \allowbreak
    \\
    \infer[\var{perm\_skip}]{\var{Permutation}~(x :: l)~(x :: l')}{%
      \var{Permutation}~l~l'
    }
    \\
    \allowbreak
    \\
    \infer[\var{perm\_swap}]{\var{Permutation}~(x :: y :: l)~(y :: x :: l)}{}
    \\
    \allowbreak
    \\
    \infer[\var{perm\_trans}]{\var{Permutation}~l~l''}{
      \var{Permutation}~l~l' & \var{Permutation}~l'~l''
    }
  \end{gather*}
\end{definition}
While this definition is useful, proofs that do induction directly on
the evidence of \var{Permutation} can sometimes get stuck on the
transitive case, hence the remark in the introduction about the
usefulness of \cref{thm:stablesort_unique}.

Before we go further, we should clarify what we mean by the word
\textit{permutation}, as it has two related meanings that are often
used interchangeably. We sometimes say that two lists are permutations
of one other if they contain all of the same elements, perhaps in a
different order, like the lists $l =
[\var{zebra},\ \allowbreak\var{appliance},\ \allowbreak\var{kite}]$
and $l' =
[\var{appliance},\ \allowbreak\var{kite},\ \allowbreak\var{zebra}]$
However, in mathematics we sometimes say that a permutation is
actually the bijection on the indices that defines the correspondence
between the two lists. In this case, we could map indices in $l$ to
indices in $l'$ by mapping $0 \mapsto 2$, mapping $1 \mapsto 0$, and mapping $2 \mapsto
0$. $p$ is the permutation that relates $l$ to $l'$ if for all indices
$i$ and $j$,
\begin{equation*}
  l_i = l'_{p(i)}
\end{equation*}
For consistency with Coq's built-in \var{Permuation} relation, in a
situation like this we will say that $l$ and $l'$ are permutations of
one other, and that $p$ is the \textit{permutation function} that
relates $l$ to $l'$.

This suggests an alternative definition for \var{Permutation}:
\begin{definition}[PermutationEx]
  \label{def:permutation_ex}
  Two lists $l$ and $l'$ are permutations of one another when there
  exists a permutation function $p$ on $n = \var{length}~l$ elements
  such that $p(l) = l'$.
\end{definition}
Here we are overloading the notation $p(l)$ to denote the result of
shuffling a list~$l$ according to a permutation function~$p$. To
define this precisely, we have to first define how we will represent
permutation functions in Coq.

@@

However, not any list of natural numbers represents a valid
permutation function, because a permutation must be a bijection. We
can define the property $\var{PermFun}~n~p$, which holds when $p :
\var{list}~\nat$ represents a permutation function on $n$ elements.
\begin{definition}[PermFun]
  A list~$p$ of natural numbers is a permutation function on $n$
  elements if for all natural numbers $i$, $i \in p \iff i < n$, and if
  $p$ contains no duplicate elements.
\end{definition}

@@

We should also mention a third definition for permutations that can be
used on types with decideable equality:
\begin{definition}[PermutationCount]
  \label{def:permutation_count}
  Two lists are permutations of one another when for any $x$,
  \begin{equation*}
    \var{count\_occ}~l~x = \var{count\_occ}~l'~x
  \end{equation*}
  where $\var{count\_occ}~l~x$ counts the number of elements in $l$
  that are equal to $x$.
\end{definition}

@@

With this we can define the specification for sorting functions:
\begin{definition}[Sort]
  \label{def:sort}
  A function $f : \var{list}~A \to \var{list}~A$ is a sorting function
  when for all inputs $l$,
  \begin{equation*}
    \var{Sorted}~(f~l) \land \var{Permutation}~(f~l)~l
  \end{equation*}
\end{definition}

@@

Now that we've defined sorts, we can consider stable sorts. Stable
sorts must preserve the relative orders of ``equivalent'' elements.
This equivalence comes from the ordering we are using: any total
decidable preorder also give us a decideable equivalence relation $\equiv$,
where $x \equiv y$ when $x \le y$ and $x \ge y$.

@@

\begin{restatable}[StablePermEx\_iff]{theorem}{stablepermexiff}
  For all lists $l$ and $l'$
  \begin{equation*}
    \var{StablePerm}~l~l' \iff \var{StablePermEx}~l~l'
  \end{equation*}
\end{restatable}

\begin{restatable}[StablePermInd\_iff]{theorem}{stablepermindiff}
  For all lists $l$ and $l'$
  \begin{equation*}
    \var{StablePerm}~l~l' \iff \var{StablePermInd}~l~l'
  \end{equation*}
\end{restatable}

@@

This can often be extended even to unstable sorting algorithms.
Remember, it isn't always true that $x \equiv y$ implies $x = y$. For
example, if we are sorting strings by their first character, then
$\text{\texttt{fit}} \equiv \text{\texttt{fog}}$, but clearly
$\text{\texttt{fit}} \neq \text{\texttt{fog}}$. When we do have that the
equivalence ($\equiv$) implies Leibniz equality ($=$), we say that we have
an \textit{order}, as we have when we are sorting strings using all of
their characters. Under an order, as opposed to a preorder, equivalent
elements are actually indistinguishable, so you can't tell if they
have been swapped.

@@

Least-significant digit (LSD) radix sort is a simple radix sort that
is used to lexicographically sort strings of equal length. We can
define what it means to sort lexicographically:

@@

In a functional setting, we can define an LSD radix sort elegantly by
repeatedly moving the last column to the front of the matrix and
sorting on the first column (\cref{fig:lsd_radixsort_func}). First, we
define \var{rrot}, which moves the last element of a list to the
front, and relies on \var{init}, which drops the last element of a
list. Note that in order to ensure that $\var{last}$ is total, it
takes an extra argument to return in the case of an empty list.
\begin{lstlisting}
Definition rrot (l: list A) : list A :=
  match l with
  | [] => []
  | hd :: tl => last l hd :: init l
  end.

@@

Here \var{hdsort} is an insertion sort
(\cref{appendix:insertion_sort}) that only examines the first
character of each string. We can implement this by defining orderings
that compare elements using some key function:

\begin{definition}[keyOrd]
  Given a preorder on a type $K$ and a key function $\var{key} : A \to
  K$, we can define a preorder on $A$ that compares all elements using
  the key function. That is, we can define the relation
  $\le_{\var{key}}$ on $A$ such that for all $x, y : A$
  \begin{equation*}
    x \le_{\var{key}} y \coloneqq (\var{key}~x) \le (\var{key}~y)
  \end{equation*}
\end{definition}

@@

\begin{definition}[lexle\_firstn]
  We define the relation $\le_n$ on $\var{list}~A$
  such that for all $xs, ys : \var{list}~A$,
  \begin{align*}
    xs \le_n ys &\coloneqq xs \le_{\var{firstn}~n} ys \\
              &= (\var{firstn}~n~xs) \lexle (\var{firstn}~n~ys)
  \end{align*}
\end{definition}
We also define $\var{PrefixSorted}~n$ as \var{Sorted} specialized to
the $\le_n$ ordering. \var{hdsort} is insertion sort specialized to the
$\le_1$ ordering.

@@

For both properties, our approach will be to prove some invariant that
holds after $j$ iterations. For \var{StablePerm}, we can say that
after $j$ iterations of rotating and sorting, the matrix should be a
stable permutation of the original matrix had we only rotated $j$
times without sorting.

@@

Let's generalize, and consider two orders $\le_j$ and $\le_k$ such that $j
\le k$. If in the first $j$ elements we can conclude that $x <_j y$,
then examining more elements can't change that.

@@

Now we can prove the stability invariant:
\radixsortstableinv
\begin{proof}
  We reason by induction on $j$. In the inductive case, we have to prove
  \begin{equation*}
    \var{StablePerm}~(\var{radixsort}~m~(j+1))~((\var{map}~\var{rrot})^{j+1}~m)
  \end{equation*}
  Expanding $\var{radixsort}~m~(j+1)$:
  \begin{align*}
     \var{radixsort}~m~(j+1)
    ={}& (\var{hdsort} \comp \var{map}~\var{rrot})^{j+1}~m \\
    ={}& \var{hdsort}~(\var{map}~\var{rrot}~(\var{hdsort} \comp \var{map}~\var{rrot})^j~m)
  \end{align*}
  By \cref{thm:stableperm_weaken}, the stability relative to $\le_1$
  that is guaranteed by \var{hdsort} implies stability relative to
  $\lexle$, so we have only to show.
  \begin{align*}
    \var{StablePerm}~&(\var{map}~\var{rrot}~(\var{hdsort} \comp \var{map}~\var{rrot})^j~m) \\
                     &((\var{map}~\var{rrot})^{j+1}~m)
  \end{align*}
  By \cref{thm:stableperm_map_rrot} (proved below), we can take off
  the outer $\var{map}~\var{rrot}$ from both sides.
  \begin{equation*}
    \var{StablePerm}~
    ((\var{hdsort} \comp \var{map}~\var{rrot})^j~m)~((\var{map}~\var{rrot})^j~m)
  \end{equation*}
  This is exactly the induction hypothesis.
\end{proof}

@@

  Here we can apply the swap ruel if we can show that $\var{rrot}~x
  \not\equivle \var{rrot}~y$. This follows from the fact that
  \begin{equation*}
    x \equivle y \iff \var{rrot}~x \equivle \var{rrot}~y \qedhere
  \end{equation*}
\end{proof}

@@

Although it is not obvious, the transformed string and the index
provide enough information to recover the the original string.
Conceptually, we can find the original string by using the transformed
string to reconstruct the entire sorted rotation matrix, and then
using the index to find the row containing the original string. This,
though, relies on our ability to reconstruct the sorted rotation
matrix from just its last column.

A key insight is that because of \cref{thm:sort_ord_unique}, we can
pretend that during the forwards transform, the rotation matrix is
being sorted using any sorting algorithm we find convenient.
Specifically, if we examine how an LSD radix sort sorts the rotation
matrix (\cref{fig:radixsort_rots}) we notice an interesting pattern.
Each column of the rotation matrix is a permutation of the original
string, so when we sort a on particular column, we always produce the
same result. As we move right-to-left, the column containing the
sorted input string moves with us until it reaches the left edge of
the matrix.

\begin{figure*}
  \centering
  \begin{tt}
  \setlength{\tabcolsep}{0pt}
  \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
    abracadabra&!& \\
    bracadabra!&a& \\
    cadabra!abr&a& \\
    dabra!abrac&a& \\
    bra!abracad&a& \\
    !abracadabr&a& \\
    racadabra!a&b& \\
    ra!abracada&b& \\
    adabra!abra&c& \\
    abra!abraca&d& \\
    acadabra!ab&r& \\
    a!abracadab&r&
  \end{tabular}
  $\rightarrow \cdots \rightarrow$
  \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
    racadabra&!&ab \\
    bracadabr&a&!a \\
    acadabra!&a&br \\
    a!abracad&a&br \\
    dabra!abr&a&ca \\
    bra!abrac&a&da \\
    cadabra!a&b&ra \\
    !abracada&b&ra \\
    abra!abra&c&ad \\
    ra!abraca&d&ab \\
    abracadab&r&a! \\
    adabra!ab&r&ac
  \end{tabular}
  $\rightarrow \cdots \rightarrow$
  \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
    a&!&abracadabr \\
    r&a&!abracadab \\
    d&a&bra!abraca \\
    !&a&bracadabra \\
    r&a&cadabra!ab \\
    c&a&dabra!abra \\
    a&b&ra!abracad \\
    a&b&racadabra! \\
    a&c&adabra!abr \\
    a&d&abra!abrac \\
    b&r&a!abracada \\
    b&r&acadabra!a
  \end{tabular}
  $\rightarrow$
  \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
    &!&abracadabra \\
    &a&!abracadabr \\
    &a&bra!abracad \\
    &a&bracadabra! \\
    &a&cadabra!abr \\
    &a&dabra!abrac \\
    &b&ra!abracada \\
    &b&racadabra!a \\
    &c&adabra!abra \\
    &d&abra!abraca \\
    &r&a!abracadab \\
    &r&acadabra!ab
  \end{tabular}
  \end{tt}
  \caption{Sorting the rotation matrix with LSD radix sort. The
    highlighted column contains the sorted original string.}
  \label{fig:radixsort_rots}
\end{figure*}

We also can notice that the columns to the right of the sorted column
move with it for similar reasons. The column directly to the right of
the shaded column is the result of sorting the input string, and then
stably sorting it again by the character that precedes it in the
original string. The next column is the result of sorting by each
character, sorting by the preceding character, and then sorting by the
character 2 spaces in front.

This suggests that we can recreate the matrix from the last column by
sorting it to produce the first column, then repeatedly prepending the
last column to the growing matrix and stably sorting by it
(\cref{fig:recreate}). As the matrix grows from right to left, we are
mimicing the process of LSD radix sort in \cref{fig:radixsort_rots},
in that the sorted portions to the right of the shaded column are
identical.

>>>>

Stable sorts are an important and fundamental type of sorting
algorithm. A sort is stable if it preserves the relative order of
equivalent elements. For example, if we are stably sorting the list
$[\var{mix},\ \allowbreak\var{fog},\ \allowbreak\var{age},\ \allowbreak\var{fit}]$
by the first character of each word, we have to produce
$[\var{age},\ \allowbreak\var{fog},\ \allowbreak\var{fit},\ \allowbreak\var{mix}]$
and not
$[\var{age},\ \allowbreak\var{fit},\ \allowbreak\var{fog},\ \allowbreak\var{mix}]$.
Even though both are sorted on the first characters, because \var{fog}
preceded \var{fit} in the original list, it must also precede
\var{fit} in the stably sorted list.

Stability can be directly useful in certain contexts, such as when
sorting a table by a single column. However, stability is also crucial
to the correctness of certain general-purpose algorithms. One such
algorithm, or set of algorithms, is radix sort. Radix sorts are used
to sort data lexicographically, and can be more efficient than
comparison-based sorting methods, both in theory and in practice
\cite{McIlroy93,Bentley:1997:FAS:314161.314321,10.1007/978-3-540-89097-3_3}.
Most radix sorts rely on a some stable sorting subroutine, as does
least-significant-digit (LSD) radix sort (\cref{sec:radix_sort}).

@@

Throughout our proofs about LSD radix sort and the BWT, we reason
about sorting algorithms whose implementation is irrelevant as long as
they are correct. We could, in principle, rely only on the fact that
they fulfill the specification of a sorting algorithm
(\cref{def:sort}). However, in practice this is difficult, because the
inductive definition of permutations can produce unhelpful induction
hypotheses, especially in the transitive case. Instead, we reason
about specifically insertion sort. By \cref{thm:stablesort_unique} we
can always replace insertion sort (which is slow) with any other
(presumably faster) stable sort.

@@

Whenever we sort a list, we do so with respect to some ordering on its
elements. For example, we could sort a list of students by first name,
last name, birthday, etc. More precisely, for sorting we need a total,
decidable preorder: a \textit{preorder} is a relation $\le$ that is
reflexive and transitive, it is \textit{total} if we can compare any
two elements, and it is \textit{decidable} if we have a procedure that
takes any two elements $x$ and $y$ and tells us if $x \le y$ or $x \ge y$.

Given a preorder, we can define when a list is in ascending order with
respect to that preorder. In this case, as will be the case with other
properties we will discuss, there are multiple reasonable definitions.
Our approach will be to state all the definitions we can think of and
to prove equivalence between them. This increases our confidence that
we have defined things correctly, and allows us to choose whichever
definition we find most convenient for a given proof.

@@

However, a sorting function cannot produce just any sorted list; the
output must also contain the same elements as the original list. To
capture this, we could use the \var{Permutation} relation defined in
the Coq standard library. This relation holds on two lists when they
contain all the same elements, and is defined inductively.
\begin{definition}[Permutation]
  \label{def:permutation}
  \begin{gather*}
    \infer[\var{perm\_nil}]{\var{Permutation}~[]~[]}{}
    \\
    \allowbreak
    \\
    \infer[\var{perm\_skip}]{\var{Permutation}~(x :: l)~(x :: l')}{%
      \var{Permutation}~l~l'
    }
    \\
    \allowbreak
    \\
    \infer[\var{perm\_swap}]{\var{Permutation}~(x :: y :: l)~(y :: x :: l)}{}
    \\
    \allowbreak
    \\
    \infer[\var{perm\_trans}]{\var{Permutation}~l~l''}{
      \var{Permutation}~l~l' & \var{Permutation}~l'~l''
    }
  \end{gather*}
\end{definition}
While this definition is useful, proofs that do induction directly on
the evidence of \var{Permutation} can sometimes get stuck on the
transitive case, hence the remark in the introduction about the
usefulness of \cref{thm:stablesort_unique}.

Before we go further, we should clarify what we mean by the word
\textit{permutation}, as it has two related meanings that are often
used interchangeably. We sometimes say that two lists are permutations
of one other if they contain all of the same elements, perhaps in a
different order, like the lists $l =
[\var{zebra},\ \allowbreak\var{appliance},\ \allowbreak\var{kite}]$
and $l' =
[\var{appliance},\ \allowbreak\var{kite},\ \allowbreak\var{zebra}]$
However, in mathematics we sometimes say that a permutation is
actually the bijection on the indices that defines the correspondence
between the two lists. In this case, we could map indices in $l$ to
indices in $l'$ by mapping $0 \mapsto 2$, mapping $1 \mapsto 0$, and mapping $2 \mapsto
0$. $p$ is the permutation that relates $l$ to $l'$ if for all indices
$i$ and $j$,
\begin{equation*}
  l_i = l'_{p(i)}
\end{equation*}
For consistency with Coq's built-in \var{Permutation} relation, in a
situation like this we will say that $l$ and $l'$ are permutations of
one other, and that $p$ is the \textit{permutation function} that
relates $l$ to $l'$. This suggests an alternative definition for
\var{Permutation}.

\begin{definition}[PermutationEx]
  \label{def:permutation_ex}
  Two lists $l$ and $l'$ are permutations of one another when there
  exists a permutation function $p$ on $n = \var{length}~l$ elements
  such that $p(l) = l'$.
\end{definition}
Here we are overloading the notation $p(l)$ to denote the result of
shuffling a list~$l$ according to a permutation function~$p$. To
define this precisely, we have to first define how we will represent
permutation functions in Coq.

@@

Not any list of natural numbers represents a valid permutation
function, because a permutation must be a bijection. We can define the
property $\var{PermFun}~n~p$, which holds when $p : \var{list}~\nat$
represents a permutation function on $n$ elements.
\begin{definition}[PermFun]
  A list~$p$ of natural numbers is a permutation function on $n$
  elements if for all natural numbers $i$, $i \in p \iff i < n$, and if
  $p$ contains no duplicate elements.
\end{definition}

@@

We should also mention a third definition for permutations that can be
used on types with decidable equality.
\begin{definition}[PermutationCount]
  \label{def:permutation_count}
  Two lists are permutations of one another when for any $x$,
  \begin{equation*}
    \var{count\_occ}~l~x = \var{count\_occ}~l'~x
  \end{equation*}
  where $\var{count\_occ}~l~x$ counts the number of elements in $l$
  that are equal to $x$.
\end{definition}

@@

With this we can define the specification for sorting functions.
\begin{definition}[Sort]
  \label{def:sort}
  A function $f : \var{list}~A \to \var{list}~A$ is a sorting function
  when for all inputs $l$,
  \begin{equation*}
    \var{Sorted}~(f~l) \land \var{Permutation}~(f~l)~l
  \end{equation*}
\end{definition}

@@

Now that we've defined sorts, we can consider stable sorts. Stable
sorts must preserve the relative orders of equivalent elements. This
equivalence comes from the ordering we are using: any total decidable
preorder also gives us a decidable equivalence relation $\equiv$, where $x
\equiv y$ when $x \le y$ and $x \ge y$.

@@

\begin{restatable}[StablePermEx\_iff]{theorem}{stablepermexiff}
  For all lists $l$ and $l'$
  \begin{equation*}
    \var{StablePerm}~l~l' \iff \var{StablePermEx}~l~l'
  \end{equation*}
\end{restatable}
\begin{proof}
  See \cref{appendix:stability}.
\end{proof}

\begin{restatable}[StablePermInd\_iff]{theorem}{stablepermindiff}
  For all lists $l$ and $l'$
  \begin{equation*}
    \var{StablePerm}~l~l' \iff \var{StablePermInd}~l~l'
  \end{equation*}
\end{restatable}
\begin{proof}
  See \cref{appendix:stability}.
\end{proof}

@@

This can be extended even to unstable sorting algorithms when $x \equiv y
\implies x = y$. This isn't always true: for example, if we are
sorting strings by their first character, then $\text{\texttt{fit}} \equiv
\text{\texttt{fog}}$, but clearly $\text{\texttt{fit}} \neq
\text{\texttt{fog}}$. When we do have that the equivalence ($\equiv$)
implies Leibniz equality ($=$), we say that we have an \textit{order},
as we have when we are sorting strings using all of their characters.
Under an order, as opposed to a preorder, equivalent elements are
actually indistinguishable, so you can't tell if they have been
swapped.

@@

Least-significant digit (LSD) radix sort is a simple radix sort that
is used to lexicographically sort strings of equal length. We can
define what it means to sort lexicographically.

@@

In a functional setting, we can define an LSD radix sort elegantly by
repeatedly moving the last column to the front of the matrix and
sorting on the first column (\cref{fig:lsd_radixsort_func}). First, we
define the right rotation of a list using \var{rrot}, which moves the
last element of a list to the front. Our definition relies on
\var{init}, which drops the last element of a list. Note that in order
to ensure that $\var{last}$ is total, it takes an extra argument to
return in the case of an empty list.
\begin{lstlisting}
Definition rrot (l: list A) : list A :=
  match l with
  | [] => []
  | hd :: tl => last l hd :: init l
  end.

@@

Here \var{hdsort} is an insertion sort
(\cref{appendix:insertion_sort}) that only examines the first
character of each string. We can implement this by defining orderings
that compare elements using some key function.

\begin{definition}[keyOrd]
  Given a preorder on a type~$K$ and a key function~$\var{key} : A \to
  K$, we define a preorder~$\le_{\var{key}}$ on $A$ that compares all elements using
  the key function.
  \begin{equation*}
    x \le_{\var{key}} y \coloneqq (\var{key}~x) \le (\var{key}~y)
  \end{equation*}
\end{definition}

@@

\begin{definition}[lexle\_firstn]
  We define the relation $\le_n$ on $\var{list}~A$
  such that for all $xs, ys : \var{list}~A$,
  \begin{align*}
    xs \le_n ys &\coloneqq xs \le_{(\var{firstn}~n)} ys \\
              &= (\var{firstn}~n~xs) \lexle (\var{firstn}~n~ys)
  \end{align*}
\end{definition}
We also define $\var{PrefixSorted}~n$ as \var{Sorted} specialized to
the $\le_n$ ordering. \var{hdsort} is insertion sort specialized to the
$\le_1$ ordering.

@@

For both properties, our approach will be to prove some invariant that
holds after $j$ iterations. For \var{StablePerm}, we can say that
after $j$ iterations of rotating and sorting, the matrix should be a
stable permutation of the original matrix had we rotated $j$ times
without sorting.

@@

In general, let's consider the relationship between two orders $\le_j$
and $\le_k$ such that $j \le k$. If when comparing two strings $x$ and
$y$, in the first $j$ elements we can conclude that $x <_j y$, then
examining more elements can't change that.

@@

Now we can prove the stability invariant.
\radixsortstableinv
\begin{proof}
  We reason by induction on $j$. In the inductive case, we have to prove
  \begin{equation*}
    \var{StablePerm}~(\var{radixsort}~m~(j+1))~((\var{map}~\var{rrot})^{j+1}~m)
  \end{equation*}
  Expanding $\var{radixsort}~m~(j+1)$:
  \begin{align*}
     \var{radixsort}~m~(j+1)
    ={}& (\var{hdsort} \comp \var{map}~\var{rrot})^{j+1}~m \\
    ={}& \var{hdsort}~(\var{map}~\var{rrot}~(\var{hdsort} \comp \var{map}~\var{rrot})^j~m)
  \end{align*}
  By \cref{thm:stableperm_weaken}, the stability relative to $\le_1$
  that is guaranteed by \var{hdsort} implies stability relative to
  $\lexle$, so we have only to show.
  \begin{align*}
    \var{StablePerm}~&(\var{map}~\var{rrot}~(\var{hdsort} \comp \var{map}~\var{rrot})^j~m) \\
                     &((\var{map}~\var{rrot})^{j+1}~m)
  \end{align*}
  By \cref{thm:stableperm_map_rrot} (proved below), we can take off
  the outer $\var{map}~\var{rrot}$ from both sides.
  \begin{equation*}
    \var{StablePerm}~
    ((\var{hdsort} \comp \var{map}~\var{rrot})^j~m)~((\var{map}~\var{rrot})^j~m)
  \end{equation*}
  This is exactly the induction hypothesis.
\end{proof}

@@

  Here we can apply the swap rule if we can show that $\var{rrot}~x
  \not\equivle \var{rrot}~y$. This follows from the fact that
  \begin{equation*}
    x \equivle y \iff \var{rrot}~x \equivle \var{rrot}~y \qedhere
  \end{equation*}
\end{proof}

@@

Although it is not obvious, the transformed string and the index
provide enough information to recover the the original string. The
crucial insight is that from the transformed string, which is the last
column of the sorted rotation matrix, we can reconstruct the entire
sorted rotation matrix. From there, we can find the row that contains
the original string using the index.

\begin{figure}
  $
  {\var{abra\textbf{c}}}%
  {\underbrace{\var{adabra\textcolor{gray}{!abrac}}}}%
  {\var{\textcolor{gray}{adabra!}}}
  $
  \caption{The last column of the sorted rotation matrix contains each
    character of the original string, sorted by the $n$ characters
    that follow it in the original string. This shows the $n$
    characters that follow \var{c} are \var{adabra!abrac}, hence that
    is ``\var{c}'s rotation''.}
  \label{fig:last_col_key}
\end{figure}

How, though, does the last column encapsulate all the information in
the sorted rotation matrix? First, we can notice that each column is a
permutation of the original string. In particular, the last column
contains each character of the original string, sorted by ``its
rotation'': the $n$ characters that follow it in the original string
(\cref{fig:last_col_key}).

We can gain more insight if we examine how the last column is created
when sorting the rotation matrix with an LSD radix sort as in
\cref{fig:radixsort_rots} (this is justified by
\cref{thm:sort_ord_unique}). After $j$ iterations, the column directly
to the left of the shaded column (the ``next column'') contains all
the characters in the original string sorted by the $j$ characters
that follow it in the original string.

\begin{figure*}
  \centering
  \begin{tt}
  \setlength{\tabcolsep}{0pt}
  \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
    abracadabra&!& \\
    bracadabra!&a& \\
    cadabra!abr&a& \\
    dabra!abrac&a& \\
    bra!abracad&a& \\
    !abracadabr&a& \\
    racadabra!a&b& \\
    ra!abracada&b& \\
    adabra!abra&c& \\
    abra!abraca&d& \\
    acadabra!ab&r& \\
    a!abracadab&r&
  \end{tabular}
  $\rightarrow \cdots \rightarrow$
  \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
    racadabra&!&ab \\
    bracadabr&a&!a \\
    acadabra!&a&br \\
    a!abracad&a&br \\
    dabra!abr&a&ca \\
    bra!abrac&a&da \\
    cadabra!a&b&ra \\
    !abracada&b&ra \\
    abra!abra&c&ad \\
    ra!abraca&d&ab \\
    abracadab&r&a! \\
    adabra!ab&r&ac
  \end{tabular}
  $\rightarrow$
  \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
    acadabra&!&abr \\
    racadabr&a&!ab \\
    cadabra!&a&bra \\
    !abracad&a&bra \\
    abra!abr&a&cad \\
    ra!abrac&a&dab \\
    abracada&b&ra! \\
    adabra!a&b&rac \\
    bra!abra&c&ada \\
    a!abraca&d&abr \\
    bracadab&r&a!a \\
    dabra!ab&r&aca
  \end{tabular}
  $\rightarrow \cdots \rightarrow$
  \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
    a&!&abracadabr \\
    r&a&!abracadab \\
    d&a&bra!abraca \\
    !&a&bracadabra \\
    r&a&cadabra!ab \\
    c&a&dabra!abra \\
    a&b&ra!abracad \\
    a&b&racadabra! \\
    a&c&adabra!abr \\
    a&d&abra!abrac \\
    b&r&a!abracada \\
    b&r&acadabra!a
  \end{tabular}
  $\rightarrow$
  \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
    &!&abracadabra \\
    &a&!abracadabr \\
    &a&bra!abracad \\
    &a&bracadabra! \\
    &a&cadabra!abr \\
    &a&dabra!abrac \\
    &b&ra!abracada \\
    &b&racadabra!a \\
    &c&adabra!abra \\
    &d&abra!abraca \\
    &r&a!abracadab \\
    &r&acadabra!ab
  \end{tabular}
  \end{tt}
  \caption{Sorting the rotation matrix with LSD radix sort. The
    highlighted column is the last column that was sorted on.}
  \label{fig:radixsort_rots}
\end{figure*}

The ``next-column'' converges to the last column as the matrix is
sorted, because on each iteration it can look one character further in
the string. The last column has full knowledge because it can look at
all $n$ characters. For example, in \cref{fig:radixsort_rots} we can
see that the ``next column'' after three iterations is
\var{ar!drcaaaabb}, which is very similar to the last column at the
end of the process, \var{ard!rcaaaabb}. The only difference is the
\var{!} and \var{d} are swapped, because the three characters that
follow them are the same (\var{abr}), so we can't tell yet which
should go first. By the end, we can see that \var{d}, whose rotation
is \var{abra!abracad}, should go before \var{!}, whose rotation is
\var{abracadabra!}.

From this perspective, the last column is the best ``next column'' for
the LSD radix sort, and we can use it at every iteration
(\cref{fig:recreate}). This gives us a way to recover the entire
sorted rotation matrix from the last column~$s$: starting with the
matrix as just a single column $s$, sort the matrix so far by the
first column, then prepend $s$ and repeat. As the matrix grows from
right to left, we are mimicking the process of LSD radix sort in
\cref{fig:radixsort_rots} in that the sorted portions to the right of
the shaded column are identical.
